---
title: "Heatmap"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(rvest)
library(knitr)
library(leaflet)
library(rgdal)
library(lubridate)
library(plotly)
col1 = "#d8e1cf" 
col2 = "#438484"
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

shooting_initial = 
  read_csv("./data/NYPD_Shooting.csv") %>% janitor::clean_names()
shooting_2021 = read_csv("./data/NYPD_shooting_New.csv") %>% janitor::clean_names()

shooting_2021 = shooting_2021 %>% 
  rename(lon_lat = new_georeferenced_column)

shooting = rbind(shooting_initial, shooting_2021) %>% 
  mutate(boro = as.factor(boro)) %>%
  mutate(location_desc = replace_na(location_desc, "NONE")) %>%
  mutate(location_desc = as.factor(location_desc)) %>%
  separate(occur_date, into = c("month", "day", "year")) %>% 
  mutate(month = as.numeric(month)) %>% 
  arrange(year, month) %>% 
  mutate(year = as.character(year)) %>% 
  mutate(boro = tolower(boro)) %>% 
  mutate(boro = if_else(boro == "staten island", "staten_island", boro)) %>% 
  rename(borough = boro) %>% 
    mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(incident_key, date, everything())
```


## Make the heatmap based on time
```{r heatmap}
shooting_heatmap = shooting_initial %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(shooting_heatmap, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap
```

According to the result of this heatmap, the midnight of weekends(Sunday and Saturday) have the highest risk of shooting cases and the total shooting cases even reach 500 for an hour. Additionally, daytime between 7 in the morning and 19 in the evening seems to have lower shooting cases than the other time of the day.

## Is this situation happened in every borough?
## Let's have a look!

### In Brooklyn
It is very similar to the heatmap above, and the total number of shooting cases is very high as well, which means Brooklyn is one of the main shooting happened place in NYC and it kind of a representative shooting incidents distribution in NYC.
```{r heatmap_Brooklyn, message = FALSE}
heatmap_bn = shooting_initial %>%
  filter(boro == "BROOKLYN") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_bn, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_Bn = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Brooklyn",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_Bn
```

### In Bronx

We can see that there is a slight difference between Brooklyn and the whole New York City, which is the total shooting cases decrease apparently. The highest case is 150 but in the total heatmap reach 500. Weekends(Sunday and Saturday) midnight still have the highest risk of shooting cases.

```{r heatmap_Bronx, message = FALSE}
heatmap_bx = shooting_initial %>%
  filter(boro == "BRONX") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_bx, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_Bx = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Bronx",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_Bx
```

### In Queens

The distribution of shooting incidents is similiar but the number of cases is below 100 in queens. Also, you can have morning jog on Wednesday and Tuesday without any stress at 6(just kiding).
```{r heatmap_Queens, message = FALSE}
heatmap_q = shooting_initial %>%
  filter(boro == "QUEENS") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_q, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_q = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Queens",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_q
```

### In Manhattan

Weekends still have higher risk and the highest total shooting cases drop to 80.

```{r heatmap_Manhattan, message = FALSE}
heatmap_m = shooting_initial %>%
  filter(boro == "MANHATTAN") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_m, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_m = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Manhattan",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_m
```


### In Staten Island 
Obviously, this is the safest borough considering shooting cases. The highest total shooting number is 25. The highest risk time is 3 to 4 a.m. on Saturday. Monday evening at 19 is high risk as well except weekends.

```{r heatmap_Staten_Island, message = FALSE}
heatmap_l = shooting_initial %>%
  filter(boro == "STATEN ISLAND") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_l, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_l = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Staten Island",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_l
```
