---
title: "P8105_final_project"
author: "Hao Zheng (hz2770), Peilin Zhou (pz2281), Xiao Ma(xm2276), Yida Wang(yw3774), Ruilian Zhang (rz2570)"
date: "11/11/2021"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)
library(rvest)
library(knitr)
library(leaflet)
# install.packages("rgdal",repos = "http://cran.us.r-project.org")
library(rgdal)
library(lubridate)
# library(dplyr)
# library(plyr)
library(plotly)
col1 = "#d8e1cf" 
col2 = "#438484"
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
Load the dataset NYPD shooting incident
```{r}
shooting_initial = 
  read_csv("./data/NYPD_Shooting.csv") %>%
  janitor::clean_names()
```
check null
```{r}
shooting_initial %>%
  summarise_all(~ sum(is.na(.)))
```
For col boro
```{r}
shooting = shooting_initial %>% 
  mutate(boro = as.factor(boro)) %>%
  mutate(location_desc = replace_na(location_desc, "NONE")) %>%
  mutate(location_desc = as.factor(location_desc)) %>%
  mutate(location_desc = fct_relevel(location_desc, c("NONE", "MULTI DWELL - PUBLIC HOUS", "MULTI DWELL - APT BUILD", "PVT HOUSE", "GROCERY/BODEGA", "BAR/NIGHT CLUB", "HOSPITAL", "COMMERCIAL BLDG", "DEPT STORE", "GAS STATION", "HOTEL/MOTEL", "SMALL MERCHANT", "RESTAURANT/DINER", "CLOTHING BOUTIQUE", "FACTORY/WAREHOUSE", "SOCIAL CLUB/POLICY LOCATI"))) %>%
  mutate(location_desc = recode(location_desc, 
                                "NONE" = "none",
                                "MULTI DWELL - PUBLIC HOUS" = "public_house", 
                                "MULTI DWELL - APT BUILD" = "apt_building", 
                                "PVT HOUSE" = "pirvate_house", 
                                "GROCERY/BODEGA" = "grocery_bodega", 
                                "BAR/NIGHT CLUB" = "bar_club",
                                "HOSPITAL" = "hospital",
                                "COMMERCIAL BLDG" = "commercial",
                                "DEPT STORE" = "dept_store",
                                "GAS_STATION" = "gas_station",
                                "HOTEL/MOTEL" = "hotel",
                                "SMALL MERCHANT" = "merchant",
                                "RESTAURANT/DINER" = "restaurant",
                                "CLOTHING BOUTIQUE" = "clothing",
                                "FACTORY/WAREHOUSE" = "factory",
                                "SOCIAL CLUB/POLICY LOCATI" = "club"
                                )) %>%
  separate(occur_date, into = c("month", "day", "year")) %>% 
  mutate(month = as.numeric(month)) %>% 
  arrange(year, month) %>% 
 # mutate(month = month.name[month]) %>% 
  mutate(year = as.character(year)) %>% 
  mutate(boro = tolower(boro)) %>% 
  mutate(boro = if_else(boro == "staten island", "staten_island", boro)) %>% 
  rename(borough = boro) %>% 
    mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(incident_key, date, everything())
```  

Next, clean the COVID-19 case count data

Importing COVID-19 case count data

```{r}
covid_counts = read.csv("./data/COVID19_data.csv", sep = ";") %>% as_tibble()
```

The clean dataset contains only day-by-day COVID-19 case count for each borough and the total case count in NYC of a particular day.
```{r}
clean_covid = covid_counts %>% 
   janitor::clean_names() %>% 
   rename(date = date_of_interest) %>% 
   select(date, contains("case_count")) %>% 
   select(-contains(c("probable_case_count", "case_count_7day_avg", "all_case_count_7day_avg"))) %>%
   separate(date, into = c("month", "day", "year")) %>% 
   mutate_all(as.character) %>% 
   mutate_if(is.character, gsub, pattern = ",", replacement = "") %>% 
   mutate_if(is.character, as.numeric) %>%
   pivot_longer(
    cols = bx_case_count:si_case_count,
    names_to = "borough",
    values_to = "borough_case_count"
  ) %>% 
  mutate(borough = gsub("_case_count", "", borough)) %>% 
  mutate(borough = recode(borough, "bx" = "bronx","bk" = "brooklyn","mn" = "manhattan","si" = "staten_island","qn" = "queens")) %>% 
  relocate(case_count, .after = borough_case_count) %>% 
  rename(total_case_count = case_count) %>% 
  mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(date, everything())

head(clean_covid)
```


## Shooting Incidents Across Time

### Distribution of shooting case across years
```{r}
shooting_year = shooting %>%
  group_by(year) %>% 
  summarise(n_obs = n())
#visualization shooting incidence trend
shooting_year %>% 
  plot_ly(x = ~year, y = ~n_obs, type = "scatter", mode = "lines++markers") %>% 
  layout(title = "shooting incidence trend through 2006 to 2020")
```


### Distribution of shooting case across months
```{r}
shooting_month = shooting %>%
  mutate(month = as.factor(month)) %>% 
  group_by(month) %>% 
  summarise(n_obs = n())

shooting_month %>% 
  plot_ly(x = ~month, y = ~n_obs, color = ~month, type = "bar") %>% 
  layout(title = "The Distribution of Shooting Incidence by month from 2006 to 2021")
```


### Distribution of shooting case across hour range in a day
```{r}
shooting_time = shooting %>% 
##format the occur_time variable to only hours.
  mutate(occur_time_hour = format(as.POSIXct(occur_time), format = "%H")) %>% 
  mutate(occur_time_hour = as.numeric(occur_time_hour)) %>% 
  group_by(occur_time_hour) %>% 
  summarise(case_numb = n())

#divide day time to 4 groups: 0-6;6-12;12-18;18-24
shooting_time = shooting_time %>% 
  mutate(occur_time_hour_type = case_when(
    occur_time_hour >= 0 & occur_time_hour < 6 ~ "0-6",
    occur_time_hour >= 6 & occur_time_hour < 12 ~ "6-12",
    occur_time_hour >= 12 & occur_time_hour < 18 ~ "12-18",
    TRUE ~ "18-24"))

ggplot(shooting_time, aes(x = occur_time_hour_type, y = case_numb, fill = occur_time_hour_type)) + geom_col(alpha = 1) + labs(x = "Occur_Time_Range",
       y = "Case Number",
       title = "Distribution of shooting case by day in NYC")

```

### Distribution of shooting case of month in 2020
### (since the covid-19 starts from March 2020, to see if there any relation between shooting and Covid.)
```{r}
shooting_2020 = shooting %>%
  filter(year == "2020") %>% 
  mutate(month = as.factor(month)) %>% 
  group_by(month) %>% 
  summarise(n_obs = n())

ggplot(shooting_2020, aes(x = month, y = n_obs, fill = month)) + geom_col(alpha = 1) + labs(x = "Month",
       y = "Case Number",
       title = "Distribution of shooting case by day in NYC")

```


## Shooting Incidents Across Space

This is an interactive map of shooting incidents from 2006 to 2021 in NYC. Incidents' details will be displayed after clicking the icon. GIS data of Boroughs' boundaries were obtained from NYC Open Data.

[Borough Boundaries](https://data.cityofnewyork.us/City-Government/Borough-Boundaries/tqmj-j8zm)
```{r}

nyc_boro = readOGR("./data/Borough_Boundaries/geo_export_2204bc6b-9c17-46ed-8a67-7245a1e15877.shp", layer = "geo_export_2204bc6b-9c17-46ed-8a67-7245a1e15877")


shooting %>% 
#  filter(year %in% c("2020", "2021")) %>%
  mutate_at(c("perp_age_group", "perp_sex", "perp_race"), funs(ifelse(is.na(.), "unknown", .))) %>% 
  mutate(labels = str_c("<b>Incident Key: </b>", incident_key, 
                    "<br>", "<b>Date: </b>", date,
                    "<br>", "<b>Borough: </b>", borough,
                    "<br>", "<b>Murdered: </b>", statistical_murder_flag,
                    "<br>", "<b>Perpetrator's Race: </b>", perp_race,
                    "<br>", "<b>Victim's Race: </b>", vic_race,
                    "<br>", "<b>Perpetrator's Age: </b>", perp_age_group,
                    "<br>", "<b>Victim's Age: </b>", vic_age_group
                    )) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addMarkers(lng = ~longitude, lat = ~latitude, popup = ~labels,
             clusterOptions = markerClusterOptions()) %>% 
  addPolygons(data = nyc_boro,
              weight = 0.85,
              label = ~nyc_boro@data$boro_name)
```


## Demographic Characteristics

Based on the shooting dataset we obtained, we could study the demographic characteristics of the perpetrators and victims to see whether there exists some potential patterns/correlations or not.

### perpetrators' characteristics
First, let's look at perpetrators' demographic information.

According to the previously checked numbers of null values in each column, there are `8295` values missing in `perp_age_group` and `8261` values missing in both `perp_sex` and `perp_race`. The information missing maybe not available or unknown at the time of the report and should be considered as either “Unknown/Not Available/Not Reported.”

```{r}
# group by perp_age_group and do a little summary
sum_perp_age = 
  shooting %>% 
  mutate(
    perp_age_group = ifelse(is.na(perp_age_group), "UNKNOWN", perp_age_group),
    perp_age_group = as.factor(perp_age_group)
  ) %>% 
  # count distinct incident case
  distinct(incident_key, .keep_all = TRUE) %>% 
  group_by(perp_age_group) %>% 
  arrange(perp_age_group) %>%
  relocate(perp_age_group) %>% 
  # filter the abnormal value
  filter(!(perp_age_group %in% c("1020", "224", "940"))) %>% 
  summarize(number = n())

# Visualize by a pie chart
sum_perp_age %>% 
  mutate(perc = scales::percent(number/sum(number))) %>% 
  ggplot(aes(x = "", y = perc, fill = perp_age_group)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = perc),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  theme_void() + 
  labs(title = "Pie chart for different perpetrator age groups") +
  theme(legend.position = "right")
```

Therefore, except those with unknown age group, most perpetrators are in the age group `18-24` and `24-44`, which takes up 39.4% in total.

```{r}
# Now look at perpetrators' sex and race
perp_sex_race = 
  shooting %>% 
  distinct(incident_key, .keep_all = TRUE) %>% 
  mutate(
    perp_sex = ifelse(is.na(perp_sex), "U", perp_sex),
    perp_race = ifelse(is.na(perp_race), "UNKNOWN", perp_race)
  ) %>% 
  mutate(perp_race = fct_infreq(as.factor(perp_race))) %>%
  group_by(perp_sex, perp_race) %>% 
  relocate(perp_sex, perp_race) %>% 
  arrange(perp_sex, perp_race) %>% 
  summarise(num = n())

# pie chart
perp_sex_race %>% 
  mutate(perc = scales::percent(num/sum(num))) %>% 
  filter(perp_sex != "U") %>%
  ggplot(aes(x = "", y = perc, fill = perp_race)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  theme_void() + 
  labs(title = "Pie chart for different perpetrator race in different sex") +
  theme(legend.position = "right") +
  facet_grid(.~perp_sex)


# Or we can try to visualize using bar chart
perp_sex_race %>% 
  ggplot(aes(x = perp_race, y = num, fill = perp_sex)) +
  geom_bar(
    stat = "identity", position=position_dodge()
  ) +
  labs(
    x = "Perpetrator's race",
    y = "Number of perpetrators",
    title = "Number of perpetrators in different races"
  ) +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "right")
```

From the `perp_sex_race` chart and the pie charts, most of the perpetrators are males (98.11%), and two races with highest rate are Black (M: 74%, F:67%) and White Hispanic(M: 13%, F: 19%).


### Victims' characteristics
There is no value missing in the victims' demographic characteristics: `vic_age_group`, `vic_sex` and `vic_race`.

What age groups are most victims in?
```{r}
vic_age_group =
  shooting %>% 
  mutate(as.factor(vic_age_group)) %>%
  group_by(vic_age_group) %>% 
  relocate(vic_age_group) %>% 
  arrange(vic_age_group) %>% 
  summarise(num = n())

vic_age_group %>% 
  ggplot(aes(x = vic_age_group, y = num)) +
  geom_bar(
    stat = "identity"
  ) +
  labs(
    x = "Victim's age group",
    y = "Number of victims",
    title = "Number of victims in different age groups"
  ) 
```

As shown above, most victims are in the age group `18-24` and `25-44`.

For victims' sex and race, similarly, first check the bar plot and pie chart.
```{r}
# The sex and race characteristics of victims
vic_sex_race = 
  shooting %>% 
  mutate(vic_race = fct_infreq(as.factor(vic_race))) %>% 
  group_by(vic_sex, vic_race) %>% 
  relocate(vic_sex, vic_race) %>% 
  arrange(vic_sex, vic_race) %>% 
  summarise(num = n())

# bar plot
vic_sex_race %>% 
  ggplot(aes(x = vic_race, y = num, fill = vic_sex)) +
  geom_bar(
    stat = "identity", position=position_dodge()
  ) +
  labs(
    x = "Victim's race",
    y = "Number of victims",
    title = "Number of victims in different races"
  ) +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "right")


# To see the percentage of different race, we can draw a pie chart
vic_sex_race %>% 
  mutate(perc = scales::percent(num/sum(num))) %>% 
  filter(vic_sex != "U") %>%
  ggplot(aes(x = "", y = perc, fill = vic_race)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = perc),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  theme_void() + 
  labs(title = "Pie chart for different victims' race in different sex") +
  theme(legend.position = "right") +
  facet_grid(.~vic_sex)

```

Most of the victims are Black (M: 71.84%, F: 68.6%), and the number of male victims in shooting cases is much more than female victims.


## Covid and shooting relations
Note: The date of shooting and covid datasets only overlap on February-December 2020, so we do the analysis base on the overlap.

```{r merge datasets and conunt incidence}
shooting_mini =
  shooting %>% 
  filter(year == "2020") %>% 
  select(c("date", "incident_key", "borough"))

shooting_covid = 
  merge(x = shooting_mini, y = clean_covid, by = c("date", "borough")) %>% 
  relocate("date", "month", "day", "year", everything()) %>% 
  group_by(date) %>% 
  add_count(borough, name = "borough_n_victim") %>% # victim number equals to the count of incident_key (includind duplicate)
  distinct() %>% 
  add_count(borough, name = "borough_n_shooting") %>% # shooting number equals to the count of distinct incident_key
  select(-incident_key) %>% 
  distinct() %>% 
  add_count(date, wt = borough_n_victim, name = "total_n_victim") %>% 
  add_count(date, wt = borough_n_shooting, name = "total_n_shooting") %>% 
  mutate(
    borough = recode(borough, 
                     "bronx" = "Bronx",
                     "brooklyn" = "Brooklyn",
                     "manhattan" = "Manhattan",
                     "queens" = "Queens",
                     "staten_island" = "Staten Island")
  )
```

First find the association between total covid cases and total shootings. 
```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of shooting cases against Covid cases within each day",
    x = "Total Covid cases",
    y = "Total shooting cases"
  )
```

See separately for each borough:
```{r}
shooting_covid %>% 
  ggplot(aes(x = borough_case_count, y = borough_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of shooting cases against Covid cases within each day",
    x = "Total Covid cases",
    y = "Total shooting cases"
  ) + 
  facet_wrap(~ borough)
```
We can see that for each borough, there is a general trend that higher number of shooting cases is associated with fewer covid cases, whereas the number of shooting cases decreases when covid case increases. This trend does not manifest in Staten Island, which may due to the limited size of data.

Then look at the association between total covid cases and total victims in shooting.
```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_victim, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of shooting victims against Covid cases within each day",
    x = "Total Covid cases",
    y = "Total shooting cases"
  )
```

We can also see it separately by borough:
```{r}
shooting_covid %>% 
  ggplot(aes(x = borough_case_count, y = borough_n_victim, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of shooting victims against Covid cases within each day",
    x = "Total Covid cases",
    y = "Total shooting cases"
  ) + 
  facet_wrap(~ borough)
```

Then explore by month.
```{r}
shooting_covid %>% 
#  mutate(month = month.name[month]) %>% 
  ggplot(aes(x = total_case_count, y = total_n_victim, group = month, color = month)) +
  geom_point() +
  labs(
    title = "Distribution of shooting victims against Covid cases within each day",
    x = "Total Covid cases",
    y = "Total shooting cases"
  ) +
  facet_wrap(~ month)
```
We can see from the above plot that there is no general pattern of the association between total shooting and total covid cases. But for the months in the middle of the year, especially in June, July and August, we can see that the number shooting cases is high when there are few covid cases, in which the inner logic is deserved to be explored.


Fit a linear regression model for total covid case with total shooting as predictor.
```{r}
total_lm = lm(total_n_shooting ~ total_case_count, data = shooting_covid) 

broom::tidy(total_lm) %>% 
  knitr::kable()

set.seed(100)
par(mfrow = c(2,2))
plot(total_lm)
```

We can see from the Residuals vs Fitted plot that the residuals is not equally distributed around the 0 horizontal line. In fact, the residuals has a pattern of decrease, indicating that the model has high goodness of fit, but violates the assumption of the normal distribution of the residual. Besides, from the Normal Q-Q plot we can see that most of the data fit the line when the theoretical quantile is under 2, indicating the data might have a normal distribution in a certain range.

Then do it with total victims as predictor.
```{r}
total_lm = lm(total_n_victim ~ total_case_count, data = shooting_covid) 

broom::tidy(total_lm) %>% 
  knitr::kable()

set.seed(100)
par(mfrow = c(2,2))
plot(total_lm)
```

### Make the heatmap based on time
```{r}
shooting_heatmap = shooting_initial %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour)

dayHour = plyr::ddply(shooting_heatmap, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total shooting cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "",
       x = "shooting cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap
```

