---
title: "P8105_final_project"
author: "Hao Zheng (hz2770), Peilin Zhou (pz2281), Xiao Ma(xm2276), Yida Wang(yw3774), Ruilian Zhang (rz2570)"
date: "11/11/2021"
output: html_document
---

```{r setup, message = FALSE}
library(tidyverse)
library(rvest)
library(knitr)
library(leaflet)
#install.packages("rgdal",repos = "http://cran.us.r-project.org")
library(rgdal)
library(lubridate)
#library(dplyr)
#library(plyr)
library(plotly)
col1 = "#d8e1cf" 
col2 = "#438484"
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
Load the historic and year-to-datedatasets of NYPD shooting incident
```{r}
shooting_initial = 
  read_csv("./data/NYPD_Shooting.csv") %>% janitor::clean_names()
shooting_2021 = read_csv("./data/NYPD_shooting_New.csv") %>% janitor::clean_names()

#A variable name in shooting_new is different from the initial data, change column name in order to merge the data frames
shooting_2021 = shooting_2021 %>% 
  rename(lon_lat = new_georeferenced_column)

shooting = rbind(shooting_initial, shooting_2021)
```
check null
```{r}
shooting %>%
  summarise_all(~ sum(is.na(.)))
```
For col boro
```{r}
shooting = shooting %>% 
  mutate(boro = as.factor(boro)) %>%
  mutate(location_desc = replace_na(location_desc, "NONE")) %>%
  mutate(location_desc = as.factor(location_desc)) %>%
  separate(occur_date, into = c("month", "day", "year")) %>% 
  mutate(month = as.numeric(month)) %>% 
  arrange(year, month) %>% 
 # mutate(month = month.name[month]) %>% 
  mutate(year = as.character(year)) %>% 
  mutate(boro = tolower(boro)) %>% 
  mutate(boro = if_else(boro == "staten island", "staten_island", boro)) %>% 
  rename(borough = boro) %>% 
    mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(incident_key, date, everything())
```  

Next, clean the COVID-19 case count data

Importing COVID-19 case count data

```{r}
covid_counts = read.csv("./data/COVID19_data.csv", sep = ";") %>% as_tibble()
```

The clean dataset contains only day-by-day COVID-19 case count for each borough and the total case count in NYC of a particular day.
```{r}
clean_covid = covid_counts %>% 
   janitor::clean_names() %>% 
   rename(date = date_of_interest) %>% 
   select(date, contains("case_count")) %>% 
   select(-contains(c("probable_case_count", "case_count_7day_avg", "all_case_count_7day_avg"))) %>%
   separate(date, into = c("month", "day", "year")) %>% 
   mutate_all(as.character) %>% 
   mutate_if(is.character, gsub, pattern = ",", replacement = "") %>% 
   mutate_if(is.character, as.numeric) %>%
   pivot_longer(
    cols = bx_case_count:si_case_count,
    names_to = "borough",
    values_to = "borough_case_count"
  ) %>% 
  mutate(borough = gsub("_case_count", "", borough)) %>% 
  mutate(borough = recode(borough, "bx" = "bronx","bk" = "brooklyn","mn" = "manhattan","si" = "staten_island","qn" = "queens")) %>% 
  relocate(case_count, .after = borough_case_count) %>% 
  rename(total_case_count = case_count) %>% 
  mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(date, everything())

head(clean_covid)
```

### Make the heatmap based on time
```{r heatmap}
shooting_heatmap = shooting_initial %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(shooting_heatmap, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap
```

According to the result of this heatmap, the midnight of weekends(Sunday and Saturday) have the highest risk of shooting cases. Additionally, daytime between 7 in the morning and 19 in the evening seems to have lower shooting cases than the other time of the day.

Is this situation happened in every boro?

In Brooklyn
```{r heatmap_Brooklyn}
heatmap_bn = shooting_initial %>%
  filter(boro == "BROOKLYN") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_bn, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_Bn = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Brooklyn",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_Bn
```
In Bronx
```{r heatmap_Bronx}
heatmap_bx = shooting_initial %>%
  filter(boro == "BRONX") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_bx, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_Bx = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Bronx",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_Bx
```

In Queens
```{r heatmap_Queens}
heatmap_q = shooting_initial %>%
  filter(boro == "QUEENS") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_q, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_q = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Queens",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_q
```

In Manhattan
```{r heatmap_Manhattan}
heatmap_m = shooting_initial %>%
  filter(boro == "MANHATTAN") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_m, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_m = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Manhattan",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_m
```


In Staten_island 
```{r heatmap_Staten_island}
heatmap_l = shooting_initial %>%
  filter(boro == "STATEN ISLAND") %>%
  mutate(occur_date = as.Date(occur_date,'%m/%d/%Y')) %>%
  mutate(occur_date = weekdays(occur_date)) %>%
  separate(occur_time, into = c("hour", "minute", "second")) %>%
  mutate(hour = as.factor(hour)) %>%
  select(incident_key, occur_date, hour) %>%
  mutate(occur_date = as.factor(occur_date),
         occur_date = fct_relevel(occur_date, "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday" , "Saturday"))

dayHour = plyr::ddply(heatmap_l, c( "hour", "occur_date"), summarise, N = length(incident_key))
attach(dayHour)

heatmap_l = ggplot(dayHour, aes(hour, occur_date)) + 
  geom_tile(aes(fill = N),colour = "white", na.rm = TRUE) +
  scale_fill_gradient(low = col1, high = col2) +  
  guides(fill = guide_legend(title = "Total Shooting Cases")) +
  theme_bw() + 
  theme_minimal() + 
  labs(title = "Time Based Heatmap in Staten_island",
       x = "Shooting Cases Per Hour", y = "Day of Week") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

heatmap_l
```


## Shooting Incidents Across Time

For the shooting incidents across time, we use three different levels to analyze. 
Firstly, we compared shooting case year by year.
### Distribution of Shooting Case of Years
```{r}
shooting_year = shooting %>%
  group_by(year) %>% 
  summarise(n_obs = n())
#visualization shooting incidence trend
shooting_year %>% 
  plot_ly( x = ~year, y = ~n_obs, type = "scatter", mode = "lines+markers") %>% 
  layout(title = "Shooting Incidence Trend from 2006 to 2020",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Frequency"))
```

By observing the data set, the shooting incidence gradually decrease from 2055 cases in 2006 to 967 cases in 2019. However, due to the Covid-19 pandemic and responses to large-scale protests over the killing of George Floyd, there is a sharply surge of shooting incidents in 2020 which have 1948 cases. Since the data for 2021 is only from January to September 30th, we are not sure whether there is a decrease in the year 2021 compared to 2020.


### Distribution of Shooting Case of Months
Then we take a look of average shooting cases between months from 2006 to 2021 in the New York City.
```{r}
shooting_month = shooting %>%
  mutate(month = as.factor(month)) %>% 
  group_by(month) %>% 
  summarise(n_obs = n())

shooting_month %>% 
  plot_ly(x = ~month, y = ~n_obs, color = ~month, type = "bar") %>% 
  layout( title = "The Distribution of Shooting Incidence by Month",
          xaxis = list(title = "Month"),
          yaxis = list(title = "Frequency"))
```

The distribution of the shooting incidence by month has a bell-shape. The shooting case concentrated in summer from May to September. The reason of this may related to the time of memorial day and the labor day.




### Distribution of Shooting Case of Time Range Across day
```{r}
shooting_time = shooting %>% 
##format the occur_time variable to only hours.
  mutate(occur_time_hour = format(as.POSIXct(occur_time), format = "%H")) %>% 
  mutate(occur_time_hour = as.numeric(occur_time_hour)) %>% 
  group_by(occur_time_hour) %>% 
  summarise(case_numb = n())

#divide day time to 4 groups: 0-6;6-12;12-18;18-24
shooting_time = shooting_time %>% 
  mutate(occur_time_range = case_when(
    occur_time_hour >= 0 & occur_time_hour < 6 ~ "0-6",
    occur_time_hour >= 6 & occur_time_hour < 12 ~ "6-12",
    occur_time_hour >= 12 & occur_time_hour < 18 ~ "12-18",
    occur_time_hour >= 18 & occur_time_hour < 24 ~ "18-24"))
shooting_time = shooting_time %>% 
  mutate(occur_time_range = factor(occur_time_range, levels = c("0-6","6-12","12-18","18-24")))

ggplot(shooting_time, aes(x = occur_time_range, y = case_numb, fill = occur_time_range)) + geom_col(alpha = 1) + labs(x = "Occur Time Range",
       y = "Frequency",
       title = "Distribution of Shooting Case by Day")

#pie chart of ratio of shooting cases in a day
ggplot(shooting_time, aes(x = "", y = case_numb, fill = occur_time_range)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_void() + 
  guides(fill = guide_legend(title = "occur Time range")) +
  labs(title = "Pie Chart for Distribution of Shooting Case by Day") +
  theme(legend.position = "right")
```

From the bar chart, it is obvious that most of the shooting cases happens in the evening and the late night, which concentrated in 18-24 and 0-6. The pie chart clearly show the occupy of shooting cases time range take place in a day.

### Distribution of shooting case across month in 2020
We would like to focus the gun violence in the year of 2020 as a critical year of surge. 
### (since the covid-19 starts from March 2020, to see if there any relation between shooting and Covid.)
```{r}
shooting_2020 = shooting %>%
  filter(year == "2020") %>% 
  mutate(month = as.factor(month)) %>% 
  group_by(month) %>% 
  summarise(n_obs = n())

ggplot(shooting_2020, aes(x = month, y = n_obs, fill = month)) + geom_col(alpha = 1) + labs(
  x = "Month",
  y = "Frequency",
  title = "Distribution of Shooting Case across month in 2020 in NYC")

```
The major rise in gun violence in the city began in 2020, after a period in which violent crime dropped to its lowest levels in more than six decades. For the first half of the year, Gun violence is relatively low as the city shut down early in the pandemic. The spike of shooting cases during June to August, which is mainly because of the death of George Floyd. 


Since 2020 is the critical year, we would like to analyze the average shooting case by month between year 2019 and 2020.
```{r}
shooting_2019_2020 = shooting %>%
  filter(year == "2019" | year == "2020") %>% 
  mutate(month = as.factor(month)) %>% 
  group_by(year, month) %>% 
  summarise(frequency = n())

ggplot(shooting_2019_2020, aes(x = month, y = frequency, fill = year)) + geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75)  + labs(
  x = "Month",
  y = "Frequency",
  title = "Shooting Case Across Month in NYC in 2019 & 2020")
```
According to the plot, besides there is year on year to growth between 2019 to 2020, the distribution of shooting case across month are the same. 


## Shooting Incidents Across Space

```{r, massage = FALSE, warning=FALSE}

shooting_map = shooting %>% 
  mutate_at(c("perp_age_group", "perp_sex", "perp_race"), funs(ifelse(is.na(.), "UNKNOWN", .))) %>% 
  mutate(labels = str_c("<b>Incident Key: </b>", incident_key, 
                    "<br>", "<b>Date: </b>", date,
                    "<br>", "<b>Borough: </b>", borough,
                    "<br>", "<b>Murdered: </b>", statistical_murder_flag,
                    "<br>", "<b>Perpetrator's Race: </b>", perp_race,
                    "<br>", "<b>Victim's Race: </b>", vic_race,
                    "<br>", "<b>Perpetrator's Age: </b>", perp_age_group,
                    "<br>", "<b>Victim's Age: </b>", vic_age_group
                    ))

nyc_boro = readOGR("./data/Borough_Boundaries/geo_export_2204bc6b-9c17-46ed-8a67-7245a1e15877.shp", layer = "geo_export_2204bc6b-9c17-46ed-8a67-7245a1e15877", verbose = FALSE)
```

```{r, message = FALSE, warning = FALSE}
polygon_color <- colorFactor(
  palette = "viridis",
  domain = as.factor(nyc_boro@data$boro_name))

shooting_map %>% 
  leaflet() %>% 
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addMarkers(lng = ~longitude, lat = ~latitude, popup = ~labels,
             clusterOptions = markerClusterOptions()) %>% 
  addPolygons(data = nyc_boro,
              weight = 0.85,
              fillColor = ~polygon_color(nyc_boro@data$boro_name),
#              fillOpacity = 0.6,
              color = "#BDBDC3",
              label = ~nyc_boro@data$boro_name)
```

This is an interactive map of shooting incidents from January 2006 to September 2021 in NYC. Incidents' details will show up after clicking the icon. Please zoom in and out the map to see the total number of shooting incidents occurred in the area in the window. Borough's name will appear when hovering over each borough.

## Map of Shooting Incidents Before and After COVID-19 in NYC

```{r, message = FALSE, warning = FALSE}
shooting_map_df = shooting_map %>% 
  filter(year == c(2018,2019,2020,2021)) %>% 
  split(.,~year)


year_map <- leaflet() %>% 
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addPolygons(data = nyc_boro,
              weight = 0.85,
              fillColor = ~polygon_color(nyc_boro@data$boro_name),
#              fillOpacity = 0.6,
              color = "#BDBDC3",
              label = ~nyc_boro@data$boro_name)

groupColors  <- colorFactor(
  palette = "viridis",
  domain = as.factor(names(shooting_map_df)))



names(shooting_map_df) %>%
  purrr::walk( function(df) {
    year_map <<- year_map %>%
      addCircleMarkers(data = shooting_map_df[[df]],
                          lng = ~longitude, 
                          lat = ~latitude,
                          popup = ~labels,
                          group = df,
                          radius = 0.3,
                          color =  ~groupColors(df)
#                          clusterOptions = markerClusterOptions(),
                ) 
  })

year_map %>%
  addLayersControl(
    overlayGroups = names(shooting_map_df),
    options = layersControlOptions(collapsed = FALSE)
  )
```

This map focuses on the shooting incidents from January 2018 to September 2021. Its purpose includes but not limited to observing the change of location and density of shooting incidents happened in the city before and after COVID-19. Year-to-year comparison can be made by selecting different years on the right upper corner.

### Map of Shooting Incidents by Perpetrator's Race/Ethicity

```{r, message = FALSE, warning = FALSE}
race_map_df = shooting_map %>% 
  filter(year == c(2019,2020,2021)) %>% 
  mutate(perp_race = tolower(perp_race)) %>% 
  split(.,~perp_race)

raceColors  <- colorFactor(
  palette = "viridis",
  domain = as.factor(names(race_map_df)))

race_map <- leaflet() %>% 
  addTiles() %>% 
  addProviderTiles("CartoDB.Positron") %>% 
  addPolygons(data = nyc_boro,
              weight = 0.85,
              fillColor = ~polygon_color(nyc_boro@data$boro_name),
#              fillOpacity = 0.6,
              color = "#BDBDC3",
              label = ~nyc_boro@data$boro_name)

names(race_map_df) %>%
  purrr::walk( function(df) {
    race_map <<- race_map %>%
      addCircleMarkers(data = race_map_df[[df]],
                          lng = ~longitude, 
                          lat = ~latitude,
                          popup = ~labels,
                          group = df,
                          radius = 0.3,
                          color =  ~raceColors(df)
#                          clusterOptions = markerClusterOptions(),
                ) 
  })

race_map %>%
  addLayersControl(
    overlayGroups = names(race_map_df),
    options = layersControlOptions(collapsed = FALSE)
  )
```

This map shows the location of shooting incidents from January 2019 to September 2021 categorized by the ethnic group of the perpetrator. Selection of Ethnic group for exploration is based on given information from the original data.

## Demographic Characteristics

Based on the shooting dataset we obtained, we could study the demographic characteristics of the perpetrators and victims to see whether there exists some potential patterns/correlations or not.

### Perpetrators' Characteristics
First, let's look at perpetrators' demographic information.

According to the previously checked numbers of null values in each column, there are `8295` values missing in `perp_age_group` and `8261` values missing in both `perp_sex` and `perp_race`. The information missing maybe not available or unknown at the time of the report and should be considered as either “Unknown/Not Available/Not Reported.”

The same shooting incident may have several victims but corresponding to the same perpetrator. Therefore, for the perpetrators' characteristics, we filter the data to make sure the incident key is unique to avoid counting the same perpetrators several times.

```{r perpetrator age group}
# group by perp_age_group and do a little summary
sum_perp_age = 
  shooting %>% 
  mutate(
    perp_age_group = ifelse(is.na(perp_age_group), "UNKNOWN", perp_age_group),
    perp_age_group = as.factor(perp_age_group)
  ) %>% 
  # count distinct incident case
  distinct(incident_key, .keep_all = TRUE) %>% 
  group_by(perp_age_group) %>% 
  arrange(perp_age_group) %>%
  relocate(perp_age_group) %>% 
  # filter the abnormal value
  filter(!(perp_age_group %in% c("1020", "224", "940"))) %>% 
  summarize(number = n())

# Visualize by a pie chart
sum_perp_age %>% 
  mutate(perc = scales::percent(number/sum(number))) %>% 
  ggplot(aes(x = "", y = perc, fill = perp_age_group)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = perc),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_void() + 
  guides(fill = guide_legend(title = "Perpetrator's Age Group")) +
  labs(title = "Pie Chart for Different Perpetrator Age Groups") +
  theme(legend.position = "right")
```

Therefore, except those with unknown age group, most perpetrators are in the age group `18-24` and `24-44`, which takes up 39.4% in total. We can see that most of them are quite young or in their middle age. The large percentage of unknown values may also because some of the perpetrators are still unknown/uncaught, which leaves a potential threat to the community.

```{r perpetrator sex and race}
# Now look at perpetrators' sex and race
perp_sex_race = 
  shooting %>% 
  distinct(incident_key, .keep_all = TRUE) %>% 
  mutate(
    perp_sex = ifelse(is.na(perp_sex), "U", perp_sex),
    perp_race = ifelse(is.na(perp_race), "UNKNOWN", perp_race)
  ) %>% 
  mutate(perp_race = fct_infreq(as.factor(perp_race))) %>%
  group_by(perp_sex, perp_race) %>% 
  relocate(perp_sex, perp_race) %>% 
  arrange(perp_sex, perp_race) %>% 
  summarise(num = n()) %>% 
  mutate(perc = round(num/sum(num), digits = 2))

# Get the position for the label of the pie charts
position1 = 
  perp_sex_race %>% 
  filter(perp_sex != "U") %>%
  mutate(csum = rev(cumsum(rev(perc*100))), 
         pos = perc*100/2 + lead(csum, 1),
         pos = if_else(is.na(pos), perc*100/2, pos))

# pie chart
perp_sex_race %>% 
  filter(perp_sex != "U") %>%
  ggplot(aes(x = "" , y = perc*100, fill = fct_inorder(perp_race))) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Pastel1") +
  ggrepel::geom_label_repel(data = position1,
                   aes(y = pos, label = paste0(perc*100, "%")),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Perpetrator's Race")) +
  theme_void() +
  labs(title = "Pie Chart For Different Perpetrator Race in Different Sex") +
  facet_grid(.~perp_sex)


# Or we can try to visualize using bar chart
perp_sex_race %>% 
  ggplot(aes(x = perp_race, y = num, fill = perp_sex)) +
  geom_bar(
    stat = "identity", position = position_dodge()
  ) +
  labs(
    x = "Perpetrator's Race",
    y = "Number of Perpetrators",
    title = "Number of Perpetrators in Different Races"
  ) +
  coord_flip() +
  guides(fill = guide_legend(title = "Perpetrator's Sex")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position = "right")
```

From the `perp_sex_race` chart and the pie charts, most of the perpetrators are males (98.11%), and two races with highest percentage are Black (M: 74%, F:67%) and White Hispanic(M: 13%, F: 19%). Two races with lowest percentage are American Indian/Alaskan Native (nearly 0% for both male and female perpetrators) and Asian/Pacific Islanders (M: 1%, F: 1%). The race information of perpetrators does not vary a lot in different sex. 


### Victims' Characteristics
Another important demographic information in shooting incidents is about the victim. What kind of characteristics do they hold? Is there some people more likely to become the victims in such tragedies?

There is only a very limited amount of missing values in the victims' demographic characteristics: `vic_age_group`, `vic_sex` and `vic_race`, which makes sense cause it's easier to gather the victims' information compared to the criminals in shootings.

What age groups are most victims in?
```{r victim age group}
vic_age_group =
  shooting %>% 
  mutate(as.factor(vic_age_group)) %>%
  group_by(vic_age_group) %>% 
  relocate(vic_age_group) %>% 
  arrange(vic_age_group) %>% 
  summarise(num = n())

vic_age_group %>% 
  mutate(perc = scales::percent(num/sum(num))) %>% 
  ggplot(aes(x = "", y = perc, fill = vic_age_group)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = perc),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_void() + 
  guides(fill = guide_legend(title = "Victim's Age Group")) +
  theme(legend.position = "right") +
  labs(
    x = "Victim's Age Group",
    y = "Number of Victims",
    title = "Number of Victims in Different Age Groups"
  )
```

As shown above, most victims are in the age group `18-24` and `25-44`.

For victims' sex and race, similarly, first check the bar plot and pie chart.
```{r victim sex and race}
# The sex and race characteristics of victims
vic_sex_race = 
  shooting %>% 
  mutate(vic_race = fct_infreq(as.factor(vic_race))) %>% 
  group_by(vic_sex, vic_race) %>% 
  relocate(vic_sex, vic_race) %>% 
  arrange(vic_sex, vic_race) %>% 
  summarise(num = n()) %>% 
  mutate(perc = round(num/sum(num), digits = 2))

# bar plot
vic_sex_race %>% 
  ggplot(aes(x = vic_race, y = num, fill = vic_sex)) +
  geom_bar(
    stat = "identity", position = position_dodge()
  ) +
  labs(
    x = "Victim's Race",
    y = "Number of Victims",
    title = "Number of Victims in Different Races"
  ) +
  coord_flip() +
  guides(fill = guide_legend(title = "Victim's Sex")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position = "right")


# To see the percentage of different race, we can draw a pie chart

# Let's get the position for the label of the pie charts first
position2 = 
  vic_sex_race %>% 
  filter(vic_sex != "U") %>%
  mutate(csum = rev(cumsum(rev(perc*100))), 
         pos = perc*100/2 + lead(csum, 1),
         pos = if_else(is.na(pos), perc*100/2, pos))

# pie chart
vic_sex_race %>% 
  filter(vic_sex != "U") %>%
  ggplot(aes(x = "" , y = perc*100, fill = fct_inorder(vic_race))) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Pastel1") +
  ggrepel::geom_label_repel(data = position2,
                   aes(y = pos, label = paste0(perc*100, "%")),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Victim's Race")) +
  theme_void() +
  labs(title = "Pie Chart for Different Victim Race in Different Sex") +
  facet_grid(.~vic_sex)

```

Most of the victims are Black (M: 72%, F: 69%), and the number of male victims in shooting cases is much more than female victims.



## COVID and shooting relations

Note: The date of shooting and COVID datasets only overlap on February-December 2020, so we do the analysis base on the overlap.

```{r merge datasets and conunt incidence}
shooting_mini =
  shooting %>% 
  filter(year == "2020") %>% 
  select(c("date", "incident_key", "borough"))

shooting_covid = 
  merge(x = shooting_mini, y = clean_covid, by = c("date", "borough")) %>% 
  relocate("date", "month", "day", "year", everything()) %>% 
  group_by(date) %>% 
  add_count(borough, name = "borough_n_victim") %>% # victim number equals to the count of incident_key (includind duplicate)
  distinct() %>% 
  add_count(borough, name = "borough_n_shooting") %>% # shooting number equals to the count of distinct incident_key
  select(-incident_key) %>% 
  distinct() %>% 
  add_count(date, wt = borough_n_victim, name = "total_n_victim") %>% 
  add_count(date, wt = borough_n_shooting, name = "total_n_shooting") %>% 
  mutate(
    borough = recode(borough, 
                     "bronx" = "Bronx",
                     "brooklyn" = "Brooklyn",
                     "manhattan" = "Manhattan",
                     "queens" = "Queens",
                     "staten_island" = "Staten Island")
  )
```


### Visualize Distribution

First we want to see if there is an underlying association between total COVID cases and total shootings. 

```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Borough")) +
  theme(legend.position = "right")
```

See separately for each borough.

```{r}
shooting_covid %>% 
  ggplot(aes(x = borough_case_count, y = borough_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) + 
  facet_wrap(~ borough) +
  guides(fill = guide_legend(title = "Borough")) +
  theme(legend.position = "right")
```

We can see that for each borough, there is a general trend that higher number of shooting cases is associated with fewer COVID cases, whereas the number of shooting cases decreases when COVID case increases. This trend does not manifest in Staten Island, which may due to the limited size of data.

Then explore by month.

```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, group = month, color = month)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  facet_wrap(~ month) +
  guides(fill = guide_legend(title = "Month")) +
  theme(legend.position = "right")
```

We can see from the above plot that there is no general pattern of the association between total shooting and total COVID cases. But for the months in the middle of the year, especially in June, July and August, we can see that the number shooting cases is high when there are few COVID cases, which is consistent with the former analysis.


Use K-means to identify clusters.

```{r fit kmeans}
kmeans_df = 
  shooting_covid %>% 
  ungroup() %>% 
  select(total_case_count, total_n_shooting)

kmeans_fit = 
  kmeans(x = kmeans_df, centers = 3)

kmeans_df %>% 
  broom::augment(kmeans_fit, .) %>%
  ggplot(aes(x = total_case_count, y = total_n_shooting, color = .cluster)) +
  geom_point() +
  labs(
    title = "K-means Clustering on Shooting Incident",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Cluster")) +
  theme(legend.position = "right")
```

```{r more clusters}
clusts =
  tibble(k = 2:4) %>%
  mutate(
    km_fit =    map(k, ~kmeans(kmeans_df, .x)),
    augmented = map(km_fit, ~broom::augment(.x, kmeans_df))
  )

clusts %>% 
  select(-km_fit) %>% 
  unnest(augmented) %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, color = .cluster)) +
  geom_point(aes(color = .cluster)) +
  facet_grid(~k) +
    labs(
    title = "K-means Clustering on Shooting Incident by Cluster Number",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Cluster")) +
  theme(legend.position = "right")
```


Fit a linear regression model for total COVID case with total shooting as predictor.

```{r}
total_lm = lm(total_n_shooting ~ total_case_count, data = shooting_covid) 

broom::tidy(total_lm) %>% 
  knitr::kable()

set.seed(100)
par(mfrow = c(2,2))
plot(total_lm)
```

We can see from the Residuals vs Fitted plot that the residuals is not equally distributed around the 0 horizontal line. In fact, the residuals has a pattern of decrease, indicating that the model has high goodness of fit, but violates the assumption of the normal distribution of the residual. Besides, from the Normal Q-Q plot we can see that most of the data fit the line when the theoretical quantile is under 2, indicating the data might have a normal distribution in a certain range.


