---
title: "Shooting and COVID"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(rvest)
library(plotly)

theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

shooting_initial = 
  read_csv("./data/NYPD_Shooting.csv") %>% janitor::clean_names()
shooting_2021 = read_csv("./data/NYPD_shooting_New.csv") %>% janitor::clean_names()

shooting_2021 = shooting_2021 %>% 
  rename(lon_lat = new_georeferenced_column)

shooting = rbind(shooting_initial, shooting_2021) %>% 
  mutate(boro = as.factor(boro)) %>%
  mutate(location_desc = replace_na(location_desc, "NONE")) %>%
  mutate(location_desc = as.factor(location_desc)) %>%
  separate(occur_date, into = c("month", "day", "year")) %>% 
  mutate(month = as.numeric(month)) %>% 
  arrange(year, month) %>% 
  mutate(year = as.character(year)) %>% 
  mutate(boro = tolower(boro)) %>% 
  mutate(boro = if_else(boro == "staten island", "staten_island", boro)) %>% 
  rename(borough = boro) %>% 
    mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(incident_key, date, everything())

covid_counts = read.csv("./data/COVID19_data.csv", sep = ";") %>% as_tibble()

clean_covid = covid_counts %>% 
   janitor::clean_names() %>% 
   rename(date = date_of_interest) %>% 
   select(date, contains("case_count")) %>% 
   select(-contains(c("probable_case_count", "case_count_7day_avg", "all_case_count_7day_avg"))) %>%
   separate(date, into = c("month", "day", "year")) %>% 
   mutate_all(as.character) %>% 
   mutate_if(is.character, gsub, pattern = ",", replacement = "") %>% 
   mutate_if(is.character, as.numeric) %>%
   pivot_longer(
    cols = bx_case_count:si_case_count,
    names_to = "borough",
    values_to = "borough_case_count"
  ) %>% 
  mutate(borough = gsub("_case_count", "", borough)) %>% 
  mutate(borough = recode(borough, "bx" = "bronx","bk" = "brooklyn","mn" = "manhattan","si" = "staten_island","qn" = "queens")) %>% 
  relocate(case_count, .after = borough_case_count) %>% 
  rename(total_case_count = case_count) %>% 
  mutate(date = str_c(month, day, year, sep = "/")) %>% 
  select(date, everything())

shooting_mini =
  shooting %>% 
  filter(year == "2020") %>% 
  select(c("date", "incident_key", "borough"))

shooting_covid = 
  merge(x = shooting_mini, y = clean_covid, by = c("date", "borough")) %>% 
  relocate("date", "month", "day", "year", everything()) %>% 
  group_by(date) %>% 
  add_count(borough, name = "borough_n_victim") %>% # victim number equals to the count of incident_key (includind duplicate)
  distinct() %>% 
  add_count(borough, name = "borough_n_shooting") %>% # shooting number equals to the count of distinct incident_key
  select(-incident_key) %>% 
  distinct() %>% 
  add_count(date, wt = borough_n_victim, name = "total_n_victim") %>% 
  add_count(date, wt = borough_n_shooting, name = "total_n_shooting") %>% 
  mutate(
    borough = recode(borough, 
                     "bronx" = "Bronx",
                     "brooklyn" = "Brooklyn",
                     "manhattan" = "Manhattan",
                     "queens" = "Queens",
                     "staten_island" = "Staten Island")
  )
```  

## Visualize Distribution

First we want to see if there is an underlying association between total COVID cases and total shootings. 

```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Borough")) +
  theme(legend.position = "right")
```

See separately for each borough.

```{r}
shooting_covid %>% 
  ggplot(aes(x = borough_case_count, y = borough_n_shooting, group = borough, color = borough)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) + 
  facet_wrap(~ borough) +
  guides(fill = guide_legend(title = "Borough")) +
  theme(legend.position = "right")
```

We can see that for each borough, there is a general trend that higher number of shooting cases is associated with fewer COVID cases, whereas the number of shooting cases decreases when COVID case increases. This trend does not manifest in Staten Island, which may due to the limited size of data.

Then explore by month.

```{r}
shooting_covid %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, group = month, color = month)) +
  geom_point() +
  labs(
    title = "Distribution of Shooting Incident Against COVID Cases Within Each Day",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  facet_wrap(~ month) +
  guides(fill = guide_legend(title = "Month")) +
  theme(legend.position = "right")
```

We can see from the above plot that there is no general pattern of the association between total shooting and total COVID cases. But for the months in the middle of the year, especially in June, July and August, we can see that the number shooting cases is high when there are few COVID cases, which is consistent with the former analysis.


## K-means Clusters

Based on the scatter plots above, we want to see whether there are specific groups within a certain range of COVID cases and shooting incidents. Therefore, we use K-means clustering to identify clusters.

```{r fit kmeans}
kmeans_df = 
  shooting_covid %>% 
  ungroup() %>% 
  select(total_case_count, total_n_shooting)

kmeans_fit = 
  kmeans(x = kmeans_df, centers = 3)

kmeans_df %>% 
  broom::augment(kmeans_fit, .) %>%
  ggplot(aes(x = total_case_count, y = total_n_shooting, color = .cluster)) +
  geom_point() +
  labs(
    title = "K-means Clustering on Shooting Incident",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Cluster")) +
  theme(legend.position = "right")
```

We can see that the data is separated into three groups, among which the first group has a small COVID case number and a wide range of shooting incidents, the second group has medium values for both metric, and the third group has the largest COVID case number as well as relatively small shooting incidents. These groups might indicate the association between shooting and COVID varies by the seriousness of COVID status.


```{r more clusters}
clusts =
  tibble(k = 2:4) %>%
  mutate(
    km_fit =    map(k, ~kmeans(kmeans_df, .x)),
    augmented = map(km_fit, ~broom::augment(.x, kmeans_df))
  )

clusts %>% 
  select(-km_fit) %>% 
  unnest(augmented) %>% 
  ggplot(aes(x = total_case_count, y = total_n_shooting, color = .cluster)) +
  geom_point(aes(color = .cluster)) +
  facet_grid(~k) +
    labs(
    title = "K-means Clustering on Shooting Incident by Cluster Number",
    x = "Total COVID Cases",
    y = "Total Shooting Incident"
  ) +
  guides(fill = guide_legend(title = "Cluster")) +
  theme(legend.position = "right")
```

The plots above shows k-means clusters with different group numbers, we can see that as the number of group increases, the range of both metrics shrinks by gradient.

## Summary

We found a negative trend of association between COVID cases and shooting incidents within each day. It is interesting that there is a rising in gun violence in Jun, July and August in spite of the small number of COVID cases. We will further explore the association by regression analysis.